---
title: "Splinets Statistical Application: mass spectrometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Splinets Statistical Application: mass spectrometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Splinets)
```


# INTRODUCTION 
An ovarian cancer tissue low-resolution SELDI-TOF mass spectrum (in orange) and its various representations as Splinet-objects in approximately 200 dimensional spline spaces: 


Reading a raw spectrum obtained from the original data set (A01.csv).

```{r}
#The file-path needs to be adequate to the location of data
# Path='../data/A01.csv'
Path='A01.csv'
Ovarian=read.table(Path,header=TRUE,sep=",",col.names=c('ms','Intensity'))
```


The original sample (orange) consisting of 15154 values. 

```{r, fig.width=8, fig.height=4}


plot(Ovarian$ms,Ovarian$Intensity,type='h',xlab='m/s',ylab='Intensity',col='orange',axes=FALSE)#,xaxt='n')
axis(1, at=seq(0, 20000, by=200),col = "blue")
axis(2, col = "blue")
```


The projection (navy-blue) to the spline space spanned on 200 equally spaced knots.

```{r, fig.width=8, fig.height=4}
#Equally spaced 200 knots,
xi1=seq(min(Ovarian$ms),max(Ovarian$ms),length.out=200)
so1 = splinet(xi1)  #Orthogonal basis of splines
OvSpl1=project(as.matrix(Ovarian),basis=so1$os) #Projection to the basis
#
plot(Ovarian$ms,Ovarian$Intensity,type='h',xlab='m/s',ylab='Intensity',col='orange',axes=FALSE)#,xaxt='n')
axis(1, at=seq(0, 20000, by=200),col = "blue")
axis(2, col = "blue")
lines(OvSpl1$sp)
```

The projection (red) to the spline space spanned on 200 non-equally spaced knots with their locations chosen by the spectrum.

```{r fig.width=8, fig.height=4}
wghts=abs(Ovarian$Intensity)/sum(abs(Ovarian$Intensity)) #Weights for knot selection
xi2=sort(sample(Ovarian$ms,200,prob=wghts)) #Random knots
         so2 = splinet(xi2) #Orthogonal basis of splines
         OvSpl2=project(as.matrix(Ovarian),basis=so2$os) #Projection to the basis
#
plot(Ovarian$ms,Ovarian$Intensity,type='h',xlab='m/s',ylab='Intensity',col='orange',axes=FALSE)#,xaxt='n')
axis(1, at=seq(0, 20000, by=200),col = "blue")
axis(2, col = "blue")
lines(OvSpl1$sp)
lines(OvSpl2$sp,col='red')
```

The projection to the splines with a sparse domain obtained by specifying five importance regions, with the graph above each of the five support intervals drawn by a different color.

```{r fig.width=8, fig.height=4}
xi3=seq(min(Ovarian$ms),max(Ovarian$ms),length.out=300)
so3 = splinet(xi3)
OvSpl3=project(as.matrix(Ovarian),basis=so3$os)

IndKn=1:length(OvSpl3$sp@knots)
Nsupp=5 #the number of support intervals
LE=c(700,6500,11400,13600,16500)
RE=c(5400,10800,12100,15900,18800)

OvSpl4=OvSpl3$sp
OvSpl4@supp=list()
for(l in 1:Nsupp){
  J=IndKn[OvSpl3$sp@knots>LE[l] & OvSpl3$sp@knots<RE[l]]
  OvSpl4@supp[[l]]=matrix(c(min(J),max(J)),nrow=1)
}

OvSpl4@der=list()
Der=sym2one(OvSpl3$sp@der[[1]])
for(l in 1:Nsupp){
  OvSpl4@der[[l]]=sym2one(Der[OvSpl4@supp[[l]][1,1]:OvSpl4@supp[[l]][1,2],],inv=TRUE)
}

OvSpl4@knots[c(OvSpl4@supp[[1]][1,1],OvSpl4@supp[[1]][1,2])]
OvSpl4@knots[c(OvSpl4@supp[[2]][1,1],OvSpl4@supp[[2]][1,2])]
OvSpl4@knots[c(OvSpl4@supp[[3]][1,1],OvSpl4@supp[[3]][1,2])]
OvSpl4@knots[c(OvSpl4@supp[[4]][1,1],OvSpl4@supp[[4]][1,2])]
OvSpl4@knots[c(OvSpl4@supp[[5]][1,1],OvSpl4@supp[[5]][1,2])]

plot(Ovarian$ms,Ovarian$Intensity,type='h',xlab='m/s',ylab='Intensity',col='grey',axes=FALSE)#,xaxt='n')
axis(1, at=seq(0, 20000, by=200),col = "blue")
axis(2, col = "blue")   
lines(OvSpl4)
```




